Blaming the Typeless: \\ Scalable, Human-Centric Python Localization

### 2018-08-23 paper/main.tex
restructure the slicing description

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{myref}
\usepackage{pgfplots}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

@@

@@

\subsubsection{Slice (Dynamic)}

\begin{figure}
\begin{lstlisting}
x = input() # Input: 42
if x < 20:
  y = 5
  z = 123
else:
  y = 0
  z = 321
assert(y != 0)
\end{lstlisting}
\begin{lstlisting}
x = 42
if x < 20:

assert(y != 0)
\end{lstlisting}
\caption{
  a) A program being sliced. An assertion gets triggered when line 8 is run. b)
  The lines included in the slice. This is not a valid python
  program, but that does not pose a problem for using membership in the slice as
  a learning feature.
}
\label{fig-slice-example}
\end{figure}

The goal of this feature, roughly equivalent in purpose to the type error
slice in Nate~\cite{learning-to-blame}, is to help eliminate terms that
cannot be the source of the crash. We compute a dynamic program slice
\cite{KOREL1988155}, the set of terms that contributed at runtime to the
observed exception, and turn this feature on for terms in the slice. We
follow the basic approach of \cite{KOREL1988155, KOREL1990187}, building a
graph of data and control dependencies. We then traverse the graph
backwards, starting at the execution step where the error occurred, to
collect the set of terms that the excepting line transitively depended on.
This excludes lines that could not have caused the exception, such as lines
that never ran, or lines that ran but had no connection to the step where
the exception happened.

\figref{slice-example} is an example of the input and output of the slicing
algorithm. The slicer takes as input a python source string and a list of inputs,
to be fed into the program whenever \lt{input()} is invoked. The slicer outputs
the list of line numbers for the lines in the slice.

The slicer would first create an execution trace of the example program, which
at each execution step includes the line number being run and the state of the
heap and variables. In the example, lines $1, 2, 3, 4, 8$ would have been run.
It then iterates forward through the trace, recording which
execution steps have a define-use relation, and which execution steps
have a test-control relation. The define-use relation is between a step where
the value of a variable is used, and the step where that variable was last
defined. In \figref{slice-example}, a define-use relationship exists
between lines 1 and 2 and between 3 and 8. A test-control relation is between
the execution of a control flow statement and any statements that were run as
a result of it. In \figref{slice-example}, a test-control relationship exists
between the if statement at line 2 and the assignments on lines 3 and 4. After
building up these relations, the slicer starts at the execution step causing an
exception (the execution of line 8), and builds the exception's set of
dependencies by iteratively adding to the set any step that has a define-use
or test-control relationship with a step already in the set. Finally, the
algorithm takes all the execution steps, determines which line number each step
was running, and then outputs this list of line numbers.

Our slicing algorithm handles every syntax node supported by the Python3 standard
library, \lt{ast}, except:
\begin{itemize}
\item Assignments where the left hand side is not a variable or a simple chain
  of attribute indexing or subscripting
\item Assignments where attribute indexing or subscripting on the left hand side
  means something other than the default, which can happen if the operations are
  overridden by a class
\item Lambda expressions
\item Set and dictionary comprehensions
\item Generator expressions
\item Await, yield, and yield from
\item Ellipsis (the token representing the tail of the argument list given to a
  varargs function)
\item Starred expressions
\item Async def functions and Asynchronous for
\item Delete statements
\item With statements
\item Raise statements
\end{itemize}

A downside of using the dynamic slice, in the case of control dependencies, is
that it excludes terms whose presence does not affect whether the exception
happened, but which could have had an effect if the term was different. In
\figref{slice-downside-example}, the conditional statement is never
executed, so the removal of the conditional would not have any effect on
the program's behavior, and it would not be included in the program slice.
The dynamic slice misses the incorrect test condition because the mistake
itself caused the conditional statement and its body to become irrelevant.

@@

>>>>

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{subcaption} 
\usepackage{booktabs}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{myref}
\usepackage{url}
\usepackage{pgfplots}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

@@

\newsavebox{\verbsavebox}

@@

\subsubsection{Slice (Dynamic)}
The goal of this feature, roughly equivalent in purpose to the type error
slice in Nate~\cite{learning-to-blame}, is to help eliminate terms that
cannot be the source of the crash. We compute a dynamic program
slice~\cite{KOREL1988155}: a set of terms that contributed at runtime to
the observed exception. This boolean feature encodes whether the term
is a member of the slice. Our slicing algorithm, which balances scalable
simplicity with coverage for beginner-written programs, is presented in
\secref{slice-algorithm}.

@@

\subsection{Dynamic Slicing Algorithm}
\label{sec-slice-algorithm}

\begin{figure}
\begin{lrbox}{\verbsavebox} 
\begin{lstlisting}[xrightmargin=0.5\linewidth]
x = input() # 42
if x < 20:
  y = 5
  z = 123
else:
  y = 0
  z = 321
assert(y != 0)
\end{lstlisting}
\end{lrbox} 
~ \hfill
\subcaptionbox{Pre-Slicing}{\usebox{\verbsavebox}}
\hfill
\begin{lrbox}{\verbsavebox}
\begin{lstlisting}[xrightmargin=0.5\linewidth]
x = 42
if x < 20:

assert(y != 0)
\end{lstlisting}
\end{lrbox} 
\subcaptionbox{Post-Slicing}{\usebox{\verbsavebox}}
\hfill
~
\caption{
  A program being sliced (left) and the resulting slice (right) with
  respect to the assertion on line 8. 
}
\label{fig-slice-example}
\end{figure}

Slicing information can prevent our model from implicating irrelevant
nodes in the fault localization. Program slicing is a well-studied
field with many explored tradeoffs (e.g., see Xu et al. for a
survey~\cite{xu2005}). We desire a slicing algorithm that can be computed
efficiently (to scale to hundreds of thousands of instances) but that will
also admit high accuracy: we achieve this by focusing on features relevant
to beginner-written programs. We
follow the basic approach of Korel and Laski~\cite{KOREL1988155,
KOREL1990187}, building a
graph of data and control dependencies. We then traverse the graph
backwards, starting at the execution step where the error occurred, to
collect the set of terms that the excepting line transitively depended on.
This excludes lines that could not have caused the exception, such as lines
that never ran, or lines that ran but had no connection to the step where
the exception happened.

\figref{slice-example} is an example of the input and output of the slicing
algorithm. The slicer takes as input a python source string and a list of inputs,
to be fed into the program whenever \lt{input()} is invoked. The slicer outputs
the program elements included in the slice.

The slicer proceeds by creating an execution trace of the example program. 
At each execution step the trace includes the line number being run and the state of the
heap and variables. In the example in \figref{slice-example}, 
lines $1, 2, 3, 4$ and $8$ are executed. 
It then iterates forward through the trace, recording which
execution steps have a define-use relation, and which execution steps
have a test-control relation. The define-use relation is between a step where
the value of a variable is used, and the step where that variable was last
defined. In \figref{slice-example}, a define-use relationship exists
between lines 1 and 2 and between 3 and 8. A test-control relation is between
the execution of a control flow statement and any statements that were run as
a result of it. In \figref{slice-example}, a test-control relationship exists
between the if statement at line 2 and the assignments on lines 3 and 4. After
building up these relations, the slicer starts at the execution step causing an
exception (the execution of line 8), and builds the exception's set of
dependencies by iteratively adding to the set any step that has a define-use
or test-control relationship with a step already in the set. Finally, the
algorithm maps all of the execution steps to their corresponding program 
element locations and returns the result.

To balance ease of prototype implementation against coverage for
beginner-written programs, our slicing algorithm handles every syntax node
supported by the Python3 standard \lt{ast} library except:
\begin{itemize}
\item Assignments where the left hand side is not a variable or a simple chain
  of attribute indexing or subscripting
\item Assignments where attribute indexing or subscripting on the left hand side
  means something other than the default, (e.g., if the operations are
  overridden by a class)
\item Lambda, generator and starred expressions
\item Set and dictionary comprehensions
\item Await, yield, and yield from
\item Variable argument ellipsis 
\item Coroutine definitions and asynchronous loops
\item Delete, with and raise statements
\end{itemize}

A downside of using a dynamic (as opposed to static) slice, in the case of
control dependencies, is that it excludes terms whose presence does not
affect whether the exception happened, but which could have had an effect
if the term was different. In
\figref{slice-downside-example}, the conditional statement is never
executed, so the removal of the conditional would not have any effect on
the program's behavior, and it would not be included in a more precise
program slice. In this example, the dynamic slice misses the incorrect
test condition because the very defect we are trying to localize
caused the conditional statement and its body to become irrelevant.
